Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-03-18 paper.qmd
finished information asymetry examples

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
abstract: In reviewing the literature of uncertainty visualisation
---

@@

## Mistakes when we misunderstand uncertainty
The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence). When uncertainty visualisaiton authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself. In these cases, the authors are more confused about what they are doing than the participants.

@@

The misunderstanding that descriptive statistics also extends to visualisation. The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature. @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.

@@

## Mathematically incomparable objects

> Edit notes: The "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point could also work here since this section is now about value extraction.

Visual inference can be seen as a process that combines multiple fields, from mathematics to psychology, to convey meaning. Unfortunately which tasks belong to mathematics, computer science, or psychology is poorly defined. The uncertainty visualisation taxonomies make the blur between these fields. A visualisation is, in a lot of ways, a statistic, or at the very least statistics are calculated prior to the visualisation being calculated, however this may be invisible to the user. A density plot does not just "appear" from the data, usually a smoothing function generates an estimated density function, which is then depicted on a plot using a line, which is then converted into information in your head using perceptual tasks and heuristics.  These overlapping but distinct fields create a high burden of spanning expertise to understand if two visualisation are equivalent in each field. This is largely caused by the concept of "information" differing dramatically between fields. According to mathematics, a sufficient statistic for a parameter contains the same amount of information as an entire sample. Therefore, if you are trying to estimate a population mean, the sample mean and the entire data set are both equivalent, however a similar statistic, such as the median, is not and will become a worse and worse approximation for the mean the more skewed your underlying distribution in. Computer science considered two pieces of information to be equivalent if they come from the same data. This definition of information is much looser and is contained within the mathematical definition of information. Thanksfully, we can turn to the grammar of graphics to try and organise and untangle the information in our graphics.

When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not. @Leland2005 depicts these steps as a pipeline, shown in @fig-pipeline. This pipeline shows every step from data observation to rendering of a graphic, but we are going to focus on the tail end of the pipeline specifically on statistics, geometry, coordinates, and aesthetics.

> Edit notes: Grammar of graphics is just a placeholder, Di suggested to replace this with my own pipeline visualisation based on some data wrangling papers.

![The grammar of graphics data analysis pipeline](grammarofgraphicspipeline.png){#fig-pipeline}

In order to visually communicate uncertainty we must:  
1) Mathematically quantify uncertainty in some way.   
2) Visually express this quantification using some geometry, axis, aesthetic, and rendering.  
3) Interpret the visual expression to extract the information using perceptual tasks.

Each of these steps have a large amount of research independent of the "uncertainty visualisation" subfield. There are numerous ways to depict uncertainty mathematically. Examples of mathematically defined objects that are used to express uncertainty inlude: confidence intervals, sampling distributions, bias, variance, prediction error, value ranges, quantiles etc. These definitions all exist in mathematics independent of the field of visualisation. Once we have some quantification of uncertainty, we need to express it visually. There are numerous options in combinations of geometry, axis, and aesthetics and modern graphical software will allow most of these combinations. There is also a large amount of research in information visualisation and psychology into how the choice of visual expression impacts a persons ability or ease with which they extract relevant information. In this sense, once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.

There are several frameworks that the uncertainty visualsiation problem is approached from. One approaches it from the concept of risk, and tries to answer how to best express uncertainty in a visualisation such that the viewer can acurately extract and interpret the risks involved. Another approach is to consider uncertainty visualsiation as a specific sub-field of high-dimensional data visualsiation, and therefore uncertainty is just another variable that needs to be expressed.

Cite gap __: represent different maths objects in visualisations
- @wu2023rational recognised it is difficult to understand when differences in performance stem from information asymmetry.

## Uncertainty as signal

This distinction is important because lower levels of the pipeline influence higher levels. Visualisation authors are almost unanimous in commenting that the information in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] *this citation has way more, check confirmation*. @Kinkeldey2014 adopts a definition from the 80s that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice versa, and computationally equivalent if that information can be extracted from both plots with similar easy and speed. This definition illustrates a similar idea to the grammar of graphics pipeline, shown in @fig-pipeline. First the mathematical information in a plot is established, and then that mathematical information is visually represented. Two plots that depict mathematical information that cannot be inferred from each other, cannot be compared on their visual elements. To prevent this requirement from being *too* limiting, when using a ground truth statistic we can consider two plots to be informationally equivalent if both visualisations depict a sufficient statistic. For example, a depiction of a sample and its average are equivalent if we are trying to infer the population mean. This does not mean they are computationally equivalent. Additionally, two plots cannot be compared on their visual elements if they contain different information as well. The reason for this is obvious. If you compare two graphics and one depicts the mean and the other depicts a median and you ask viewers to extract the median value, the graphic that *actually* depicts that statistic will perform better. People cannot percieve information that is not depicted.

@@

# Noise and Signal

@@

Another method that has been used is to visualise samples from the distribution we are trying to draw inference on instead of estimates along with a variance or error. This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] . The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html). Even though @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result. Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated. Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments. This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing. These sampling methods can show the messiness of the data that sits behind a model.

@@

The concept of uncertainty should also be formalised within the grammar of graphics. @Leland2005 referred to uncertainty as "semantics", however the depth of integration required to understand problems in uncertainty visualisation suggest they *need* to be formally placed within the framework. Displaying a distribution is not as simple as displaying the data. For example, the distinction between a sample and a smothed density function is not cut and dry. One may view them both as the same data but one is represented as a point, the other is binned according to some function and then represented as a line. However, this distinction blurs as n gets larger and larger and, depending on the collision modifier used, depictions of the sample will appear identical to depictions of the mass. This is obviously true due to large sample theory, the backbone of statistics. The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here. Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these veriables differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear. The lack of formalisation in this way allows authors to compare graphics that *seem* to be similar mathematically but are actually conveying very different information. Not only would a formalisation of uncertainty visualisation within the grammar of graphics allow us to iron out some of these confusions, but they would make it easier to understand when existing visualisation methods apply and can be used to explain our results.

>>>>

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
abstract: In reviewing the literature of uncertainty visualisation
---
# Current Todo List
This is a list of tasks that need to be completed before I consider the paper complete. Please keep this in mind when reading.

@@

## Mistakes made when we misunderstand uncertainty
The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence). When uncertainty visualisaiton authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself. In these cases, the authors are more confused about what they are doing than the participants.

@@

The misunderstanding that descriptive statistics also extends to visualisation. The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature. @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty. Exploratory visualisations often differ from descriptive statistics because the explicit statistic we are drawing inference on is less explicit. A versatile visualisation such as a scatter plot allows for a viewer to consider several hypothesis at once, each of which may have its own associated uncertainty. @Hullman2021 argue that there is no such thing as a "model-free" visualisation, therefore visualisation require robust visualisations of uncertainty as we are always performing inference. While we agree with this sentiment, it is clear that uncertainty *cannot* be defined without a specific motivating question, and therefore trying to include uncertainty in an exploratory visualisation stage (which by definition does not have a hypothesis) is not possible.

@@

## Mistakes made when we misunderstand uncertainty visualisation
In the previous section, we highlighted several problematic experimental designs or pieces of research that were the result of misunderstandings around the definition of uncertainty. Here, we will discuss the issues that arise due to a misunderstanding of "uncertainty visualisation". The two primary problems are the regular comparisons of mathematically incomparable plots, and reproducing results that are already widely understood in the larger information visualisation literature.

### Information asymetry in evaluated plots
The erroneous notion that "uncertainty visualisations" are a class of visualisations that depict "uncertainty" (something that is undefined) has allowed authors to compare plots that depict wildly different information with little consideration of the value of this practice.

Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12*). However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is. This is easily illustrated by the wide array of comments made by visualisation authors when they explain why two representations are being compared in their evaluation study. @Ibrekk1987 displayed 6 PDFs in their experiment because "formally equivalent representations are often not psychologically equivalent", but do not explain why what makes a representation "formally equivalent" or why only one representation was presented for the 95% CI, box plot, and CDF. @Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two. *Other examples* (*Cite Gap 13*). @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed. It is clear this lacking definition of "information" in a visual is reminicent of our problems with the definition of "uncertainty".

Unlike uncertainty, the concept of the "information" in a plot is not entirely avoided and works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot. When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not, this pipeline is illustrated in @fig-pipeline, and contains all the steps of the grammar of graphics [@Leland2005]. First the mathematical information in a plot is established in the "varset" setps, and then that mathematical information is visually represented, which is defined in the "graph" steps. Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical. However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage. Additionally, the grammar of graphics was not designed to extend to "uncertainty visualisations" as @Leland2005 referred to uncertainty as "semantics", so concepts that are common to uncertainty mathematically, such as resampling do not quite fit within the framework. For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line. These two plots differ in *both* the statistic and geometry stages of the grammar of graphics, however as n (the number of observations in our sample) gets larger, these two plots will become visually indistinguishable (depending on the collision modifier used). This result aligns nicely with large sample theory, because as n increases the data and mass should become mathematically indistinguishable, and many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness so this result is likely also practically true [@Kale2018; @Newburger2022; @Hofmann2012] but despite the increasing similarity in what is actually depicted in the graphic, the sample and mass visualisations would remain different visualisations within the grammar of graphics framework. Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable. @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result. The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here. Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these graphics differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.

![The grammar of graphics data analysis pipeline from @Leland2005. the "varset" stages are the mathematical components of the graphic, the mathematical components define the "information" in the graph, while the "graph" stages are the visual elements that determine the "computational" cost of the graph.](grammarofgraphicspipeline.png){#fig-pipeline}

Alternative information consideration within well defined frameworks may also be worthwhile to consider. @wu2023rational tried to eliminate information asymetry by esentially looking at a visualisation and identifying what information can be extracted and then using a "rational agent benchmark" to determine how a rational person would use that information for a decision making task. While this solution is interesting, it may have different results depending on who is applying it as they may use different methods or notice different elements when extracting information from a graphic. The mathematically defined concept of sufficient statistics is also scarcely mentioned likely because it is specific to a particular ground truth statistic, however uncertainty *is* specific to a particular statistic, so it could be a useful avenue for understanding when two plots differ in their ability to answer a particular question.

Ultimately, this all points to confusion as to when two visualisation contain information asymmetry, a problem that becomes more pronounced in uncertainty visualisation. @Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on the basis that all of them were "uncertainty visualisations". They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot). @Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean. They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution. @Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted a single outcome of the joint distribution of $A$ and $B$ in each frame of an animation, a comparison that was justified because all plots were "uncertainty visualisations". They found that the plot that depicted outcomes from $P(A, B)$ where the distance between the two outcomes was equivalent to $a-b$ was better at answering the question "What is $P(B<A)$" than violin or error bar plots depicting $P(A)$ and $P(B)$ separately. *Other examples* (*Cite Gap 14*).

This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymetry are regularly compared because they are both "uncertainty visualisations". This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results. This issue likely results in the lacking mathematical definition preventing us from defining and quantifying "uncertainty" mathematically and then including this quantification in a well defined framework such as the grammar of graphics.

### Repeating perceptual task experiments
Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical. Unfortunately this still often results in uninteresting or

Once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.

@@

# Uncertainty as noise and Uncertainty as signal

@@

Another method that has been used is to visualise samples from the distribution we are trying to draw inference on instead of estimates along with a variance or error. This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] . The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).  Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated. Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments. This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing. These sampling methods can show the messiness of the data that sits behind a model.

@@

The concept of uncertainty should also be formalised within the grammar of graphics. Not only would a formalisation of uncertainty visualisation within the grammar of graphics allow us to iron out some of these confusions, but they would make it easier to understand when existing visualisation methods apply and can be used to explain our results.
